Log Journal Project Theme 9

### 2022-11-09 Paper.Rmd
Added the project proposal

5. [References](#references) \dotfill 13

@@

Protein activities and their subcellular locations in cells are strongly connected. Finding proteins' positions within cells is frequently the initial step in understanding them, and their sites serve as a model for medication design. The amino acid sequence of a protein mostly determines where it will localize within a cell. Using this knowledge, a software has been developed to define a simple model of classification of yeast localization in cells. The efficient arrangement of biological data into databases and the creation of algorithms, such string algorithms, to manipulate that data, have largely defined the area of computational biology. The dataset consists of 1484 yeast sequences from SWISS-PROT using the annotations from YPD.

Research question: Can the cellular localization Sites of proteins be predicted using probabilistic classification.

@@

It is always important to look at the data and to look at the quality of the data. No proper conclusion can be drawn from low quality data. For recurrent errors the data can be corrected. The data needs to be modified to make sure the data gets the highest quality to perform the analysis on. This is achieved by performing exploratory data analysis on the dataset. Modifying and removing parts of the dataset to get best. A RandomForest classifier has been chosen for the classification of the unknown instances. The parameters has been optimized for usability and speed. See the [logjournal](https://github.com/AlfonsoJan/Theme9).

A java application has been developed. The software using Weka carried out the calculations required to carry out the probabilistic inference. The application can quickly classify new instances given from a file on from the command line. See the [repo](https://bitbucket.org/janalfonso/wekawrapper/) for the code.

@@

To create a great dataset, changes need to be made to the downloaded dataset. The dataset consists of 1484 yeast sequences from SWISS-PROT using the annotations from YPD. In this section, we are going to talk about the results we found from the dataset.

@@

Eight features were used in classification: the presence or absence of an HDEL pattern as a signal for retention in the endoplasmic reticulum lumen (erl); The results of discriminant analysis on the amino acid content of vacuolar and extracellular proteins (vac); the result of discriminant analysis on the amino acid composition of the 20-residue N-terminal region of mitochondrial and non-mitochondrial proteins (mit); the presence or absence of nuclear localization consensus patterns combined with a term reflecting the frequency of basic residues (nuc); and some combination of the presence of a short sequence motif and the result of discriminant analysis of the amino acid composition of the protein sequence (pox).

@@

Weka was used to test which algorithm is the best to use for the classification of instances. An accurate algorithm is needed that accurately predicts the dataset. It needs a high percentage of correct classifications on the dataset. An algorithm needs to be fast for the best usage. What if there is a dataset of 1 million observations. And it should also not contain too many false negatives and false positives.

@@

An integrated measurement of a measurably effect or phenomena is the area under the curve. A higher Area under the curve the better. RandomForest has the highest area under the curve, while OneR has the lowest with 0.49 and ZeroR with a close second last with 0.5.

@@

RandomForest also has the highest average precision with 0.66. And OneR has the lowest with 0.04. ZeroR is not in this plots because it only consist of NaN.

Several algorithms cannot make a prediction for the classification VAC (0.000) and for some they get an error (?). 1 algorithm IBk has a very low prediction for VAC. All the algorithm above in the figure except for ZeroR, OneR, SMO and SimpleLogistic were used in the Weka Experimenter to compare.

@@

The research question is: can machine learning predict protein localization based on their aminoacid sequences? And to answer this question properly, a clean dataset is needed to create a model. The results will show what is wrong with the data and the correlations between the different classifications. The results are not completely trustworthy because there is too little of them. Namely, there are only 1484 oberservations. This is too few for prediction with 10 different classifications.

@@

In the case of the Pox column, something strange was noticed. Out of 1484 datapoints, 15 are not zero. 11 of them have the classification of Pox. This means that if the value is not zero, the classification is highly probable Pox.

Only 8 classifiers had been used. From our research, we found that RandomForest is the best, however, perhaps another untested classifier is much better than RandomForest. Only the removal of VAC was considered what if much better precision and classification would have run if another classifier had been removed.

@@

The aim was to create a clean dataset to be used for our machine learning project. There is a clean dataset with no missing values and the correct data types. The data gives a clear overview of the protein classification and their attributes. However, not much was available. With only 1484 observations, this is just too little. And there is a very big difference between the number of observations per classification. For ERL there are only 5 observations, and for CYT there are 463. There is too big a difference between these. By improving this dataset over time, the prediction can be perfected even more. These constant improvements are important to find.

@@

@@

>>>>

5. [Project Proposal](#project-proposal) \dotfill 13

6. [References](#references) \dotfill 14

@@

Protein activities and their subcellular locations in cells are strongly connected. Finding proteins’ positions within cells is frequently the initial step in understanding them, and their sites serve as a model for medication design. The amino acid sequence of a protein mostly determines where it will localize within a cell. Using this knowledge, a software program has been developed to define a simple model for the classification of yeast localization in cells. The efficient arrangement of biological data into databases and the creation of algorithms, such as string algorithms, to manipulate that data have largely defined the area of computational biology. The dataset consists of 1484 yeast sequences from SWISS-PROT using the annotations from YPD.

Research question: How accurate can the cellular localization sites of proteins be predicted using probabilistic classification?

@@

It is always important to look at the data and to look at the quality of the data. No proper conclusion can be drawn from low-quality data. For recurrent errors, the data can be corrected. The data needs to be modified to make sure it is of the highest quality to perform the analysis on. This is achieved by performing exploratory data analysis on the dataset. Modifying and removing parts of the dataset to get the best possible dataset. A RandomForest classifier has been chosen for the classification of the unknown instances. The parameters have been optimized for usability and speed. See the [logjournal](https://github.com/AlfonsoJan/Theme9) [1].

A Java application has been developed. The software using Weka carried out the calculations required to carry out the probabilistic inference. The application can quickly classify new instances given from a file or from the command line. See the [repo](https://bitbucket.org/janalfonso/wekawrapper/) [2] for the code.

@@

To create a great dataset, changes need to be made to the downloaded dataset. The dataset consists of 1484 yeast sequences from SWISS-PROT using the annotations from YPD. In this section, we are going to talk about the results we found from the dataset [3].

@@

Eight features were used in classification: the presence or absence of an HDEL pattern as a signal for retention in the endoplasmic reticulum lumen (erl); The results of discriminant analysis on the amino acid content of vacuolar and extracellular proteins (vac); the result of discriminant analysis on the amino acid composition of the 20-residue N-terminal region of mitochondrial and non-mitochondrial proteins (mit); the presence or absence of nuclear localization consensus patterns combined with a term reflecting the frequency of basic residues (nuc); and some combination of the presence of a short sequence motif and the result of discriminant analysis of the amino acid composition of the protein sequence (pox) [3].

@@

Weka was used to test which algorithm is best to use for the classification of instances. An accurate algorithm is needed that accurately predicts the dataset. It needs a high percentage of correct classifications on the dataset. An algorithm needs to be fast for the best usage. What if there is a dataset of 1 million observations? And it should also not contain too many false negatives or false positives.

@@

The area under the curve is a way to measure an effect or phenomenon as a whole. The greater the area under the curve, the better. RandomForest has the largest area under the curve, while OneR has the lowest with 0.49 and ZeroR is a close second with 0.5.

@@

RandomForest also has the highest average precision, at 0.66. OneR has the lowest, at 0.04. ZeroR is not in this plot because it only consists of NaN.

Several algorithms cannot make a prediction for the classification VAC (0.000), and for some they get an error (?). The IBk algorithm has a very low prediction for VAC. Except for ZeroR, OneR, SMO, and SimpleLogistic, all of the algorithms in the figure above were compared in the Weka experimenter.

@@

The research question is: can machine learning predict protein localization based on their amino acid sequences? And to answer this question properly, a clean dataset is needed to create a model. The results will show what is wrong with the data and the correlations between the different classifications. The results are not completely trustworthy because there are too few of them. Namely, there are only 1484 oberservations. With ten different classifications, this is insufficient for prediction.

@@

In the case of the Pox column, something strange was noticed. Out of 1484 data points, 15 are not zero. 11 of them have the classification of Pox. This means that if the value is not zero, the classification is highly probable Pox.

Only eight classifiers had been used. From our research, we found that RandomForest is the best, however, perhaps another untested classifier is much better than RandomForest. Only the removal of VAC was considered, even though much better precision and classification would have been achieved if another classifier had been removed.

@@

The aim was to create a clean dataset to be used for our machine learning project. There is a clean dataset with no missing values and the correct data types. The data provides a clear picture of protein classification and attributes. However, not much was available. With only 1484 observations, this is just too little. And there is a very big difference between the number of observations per classification. For ERL there are only 5 observations, and for CYT there are 463. There is too big a difference between these. By improving this dataset over time, the prediction can be perfected even more. These constant improvements are important to find.

@@

# Project Proposal

I will be taking the minor in Application Design. And we have been using the program Weka this quarter, and when you open the program, you immediately see how clustered and old-fashioned this program is. My project proposal is to design a modern Weka interface.

@@

[1] = J.A. Busker, Theme 9, (2022), GitHub repository, https://github.com/AlfonsoJan/Theme9

[2] = J.A. Busker, Weka Wrapper, (2022), BitBucket repository, https://bitbucket.org/janalfonso/wekawrapper

[3] = P. Horton, K. Nakai, A Probabilistic Classification System for Predicting the Cellular Localization Sites of Proteins, (1996), AAAI, https://www.aaai.org/Papers/ISMB/1996/ISMB96-012.pdf
